Modeling Human Interactions in Collaborative Interactive Evolutionary Computation

### 2017-02-04 2017-GECCO-Modeling-User-Interactions-in-IEC/paper-body.tex
More corrections

Some C-IEC systems promote user engagement by presenting interesting information to 
users, for instance the genetic lineage of each phenotype or the most popular or 
best rated solucion\cite{picbreeder,forms}.In a recent work by 
Wagy \& Bongard \cite{wagy2014collective} user interaction 
is needed for evaluating the fitness and developing
new designs of robot locomotion. Collaboration is encouraged by gamifying the system 
using the maximum distance indicator to inspire the user to try and ``beat'' previous designs. 
In any case, using gamification techniques imply that we must deal with interactive evolutionary
system as socio-technical constructs, where the social aspects are
essential to understand its dynamics. In this sense, conclusions
reached with other systems such as NodEO \cite{DBLP:conf/gecco/MereloCGCRV16}
can also be applied to this system; and applying social 
network techniques such as graph analysis
to their study will allow us to understand them more thoroughly. 
% C-IEAS
Given the human fatigue limitation when applying IEAs, some authors 
have tried to mitigate the problem by allowing the algorithm to 
collaborate with the user, so that sometimes 
users perform the evaluation,  but also specific measures are included 
into the algorithm to perform the
evaluation of some features automatically. For instance, 
Reis et al. added some terrain measures (such as accessibility and edge length) 
to a standard  IEC in \cite{DBLP:journals/soco/FradeVC12}. 
This way the algorithm was capable of providing  terrains that would otherwise have needed
users evaluation for these specific features. Seyama and Munetomo \cite{seyama2016development}
also propose the reduction of the user fatigue by using 
a collaborative filtering algorithm to show only the information utilized by similar users as 
they collaborate with a large number of users for the interactive modeling of 3D glasses. 
The framework and data model for 
representing the interactive process of C-IEC systems is proposed next
% Volunteer Based
% HBC
% Engagement Techniques
% Models

@@

\section{Human-Centered IEC Framework}
\label{sec:framework}
The general goal of this line of research is to develop a human-centered \cite{gasson2003human} 
software framework for interactive evolutionary computation (IEC). 
A framework is defined as a reusable architectural design together
with an implementation \cite{campbell1991choices}, in this case 
providing generalized components to developers of IEC systems.

The proposed framework includes components that can be refined to increase
participation and also to minimize the amount of evaluations needed for the evolutionary 
process in a given IEC application. Software frameworks often have a vison or a way 
of thinking \cite{carneiro2010introducing} guiding their
design. Before diving into details the main design considerations of the
framework are explained next:

\begin{itemize}
\item {\bf Users are human} 
  The framework follows the approach of human centered computing \cite{sebe2010human},
  in which the context, environment, interfaces, preferences, accessibility, human relations,
  cognitive limitations, culture, creativity and other human aspects are an integral part 
  of the system. Humans are the computing resorces of the system, having unique characteristics
  as those identified by Sun \& Dance \cite{Sun2013}
\begin{quote}
  (1) humans can solve computer hard problems; (2) humans are very good at exception handling,
  (3) humans have creativity, (4) humans have cognitive load limitation, (5) humans are
  vulnerable to psychological manipulation, (6) humans are prone to errors,
  especially for reflective tasks.
\end{quote}

\item {\bf Users are volunteers} Users donate their computing resources, so they are 
unaccountable, some times they try to game the system. Project owners must actively promote and
design the interactive system to engage volunteers \cite{oh2015clicking}. % Define
\item {\bf Users are not alone}
  Relationship between users in an interactive evolutionary algorithm can be modeled
  as a social network, with well established semantics, algorithms and metrics 
  \cite{ahuja1993network}.
  A graph model could enable researchers to find other ways of identifying leaders of 
  opinion or measuring the similarity between user's preferences. 
  These measures can then be used by recommender algorithms selecting 
  phenotypes according users' preferences.

\item {\bf Context of use matters}
  \textit{Context} In computer science Fischer \cite{fischer2012context}
  defines context as
  \begin{quote}
  the interaction between humans and
  computers in socio-technical systems that takes place in a certain
  context referring to the physical and social situation in which
  computational devices and environments are embedded.
\end{quote}   
  Fischer also identifies the important aspects to consider when the context is used: how it is
  the contextual data obtained, how the context is represented and what
  goals and purposes the context has when is used in a particular
  application.   An IEC system will  be used within a certain range 
  of technical, physical and social or
  organizational environments \cite{maguire2001context} that may affect its use.

\item {\bf Interaction is a stream of actions}
  Real time processing of users' actions could be needed for certain applications when data is 
  captured by sensors, or directly captured as user input. For example, social networks keep
  track of users' interactions with other users, media objects and places. Users of 
  social networks (for instance the Facebook Graph) are accustomed to express these 
  complex relationships in sentences such as: ``John and Ann eating breakfast at Tony's''. 
  Other example is the W3C Activity Streams 2.0 specification used for representing activities 
  common in social web applications \cite{json:streams}. 
\end{itemize}

@@

\begin{itemize}
  \item {\bf Interactive Evolutionary Computational system} 
  This is the real world system that we are going to represent in the data model, 
  it consist of human users and their interactions with one or more phenotypes
  from the population. There are many ways in which humans could interact 
  with phenotypes. The interaction consists of a set of actions and 
  takes place in a certain context. For example, through a mobile device  
  or by interacting with real world objects 
  \cite{de2014artists,de2013unplugging}. 
  There is also the possibility that fitness or even part of the search 
  is done by devices lent by humans \cite{DBLP:conf/gecco/MereloCGCRV16}.
  The information gathered through the interaction is the primary concern
  as it will guide the search.

  \item {\bf IEC Data Model}
  A data model is used to describe the IEC system prior to a physical 
  implementation.  Depending on the domain several techniques can be used.
  Parts of the system could be better described by an Entity relationship 
  modeling to be used in a relational database or a class digram for a 
  key-value store. A graph is proposed for modeling the social network of users 
  and their interaction with the population. When implemented a graph database 
  will be the back-end of the system.

  \item {\bf Evolutionary Algorithm} 
  The EA algorithm interacts with the data model. The EA could also be run with the help of
  humans, for instance in the XY project \cite{de2013unplugging} artists actually painted
  the solutions of the new population and applied every genetic operator.
\end{itemize}

@@

\begin{itemize}
\item {\bf EvoSpace-Redis} 
EvoSpace is a population store \cite{Evospace}  for the development of 
evolutionary algorithms that are intended to run on a cloud computing model. 
The population is decoupled from any particular evolutionary algorithm. 
Candidate solutions are stored as of objects in a population, and they can be withdrawn, 
processed and replaced using a specified set of methods \cite{GValdez2015}. The population
is stored in-memory, using the Redis key-value database. Redis was chosen over a relational 
database management system, because it provides a hash based implementation of sets and 
queues which are natural data structures for the EvoSpace model. Basicaly a sample of 
candidate solutions are retrieved from the server, evaluated, and then sent back. 
The same operations are used to Evolve the population. In EvoSpace individuals replaced 
in the population are stored indefinitly, to permit users to store permalinks to them.
Implementation details are presented in \cite{garcia2013evospace}.

\item {\bf Graph-Neo4J} 
 Collaborative IEC (C-IEC) systems need to store highly connected data, as it is common 
 in current applications like social networks. In order to deal with large datasets of connected 
 data found in these systems, graph databases \cite{angles2012comparison} have been proposed 
 as an alternative to relational databases which have performance limitations when dealing with 
 highly connected data \cite{holzschuher2013performance}.
 A graph is proposed for modeling the social network of users, their interaction with 
 candidate solutions, and the relationships between them in the population.
 The graph database system used in the implementation is Neo4J as it is
 a scalable solution \cite{miller2013graph,holzschuher2013performance}, and well 
 supported and documented in PaaS infrastructures like Heroku. The Cypher query 
 language it used to retrieve views from the graph.
 An example query is shown in figure \ref{fig:cypher} where the relations 
 between users and solutions are presented.

@@

\item {\bf Relational Objects-PostgreSQL}
  The PostgreSQL relational database system is also employed because user 
  sessions and authentication,
  as well as dynamic web pages are handled directly by the Django web
  framework \cite{garcia2013evospace}. % these implementations details
                                % could go if needed.

\item {\bf Interfaces}
The appropiate human interface will depend on the application domain, 
the current framework implemantation employs a web based application. Developers 
can use different templates, to present one or more phenotypes 
at a time, and the type of rating system: like or one to five stars. 
Another implementations do not need a graphical interface at all, for example the
EvoDrawings instalation or the XYZ Project \cite{de2014artists}.

\item {\bf Evolution}    
The evolutionary algorithm is decoupled from the data model, the algorithm 
could be implemented as an EvoWorker \cite{garcia2013evospace}, use an external library, 
or even assigned to humans. An additional tool was implemented for a human based evolution, 
where volunteers select the parents of the next generation and upload the 
new individuals \cite{de2014artists}.  
\end{itemize}

\section{Case Study: EvoDrawings}
\label{sec:evodraw}
As a case study, a IEA was developed using the 
EvoSpace-Interactive (ES-I) platform \cite{garcia2013evospace}
A brief description of the application is presented next, focusing
on the data elements that where ported to the graph model.

@@

\subsection{Fitness Assignment}
The ES-I platform has been programmed to employ a 
collaborative methodology for performing fitness assignment. 
Several registered users assign a quality assessment to a single
phenotype and then an aggregated fitness value is calculated,
depending on the rating (from 1 to 5 stars) by each particular user,  
resulting in a many-to-many relationship between users and
phenotypes. Other components or users of the systems can 
query this user-phenotype relation to extract relevant
knowledge about the process and the population, for instance showing the
most popular phenotype, or the user with more participation 
\cite{picbreeder}.
In order to do this, metadata about phenotypes 
must be permanently stored, even
if they are no longer in the active population.

\subsection{Collaboration}
\label{sec:col}
Users need to authenticate themselves to the system using their
Facebook account. In that sense, some users might not be interested
either because they do not want to give that information or simply because
they do not use that particular social network. Even as we might lose
some users that way, the additional information we obtain for scoring
phenotypes more than compensates for that. 
After entering the web application by using their Facebook account,
users can collaborate with their Facebook friends, 
sharing those phenotypes they like, or by taking phenotypes
from their friend's collections by using the web interface depicted 
in Figure \ref{fig:web}.
At the top left corner a list of Facebook friends is presented
to encourage users to interact with the system. In the central 
\emph{ Wall } area, a phenotype sampled from the population that is
being evolved via the evolutionary algorithm 
is shown to the user.
Here, the user can interact with the system in two ways.
First, he can assign a rating to the phenotype using
a five star rating system or,
additionally, a user can choose to add an image to one of their \emph{Collections}.
A collection is a special folder that stores those phenotypes a user likes and wishes
to save. After the user finishes interacting with the phenotype
on the Wall, he can choose to retrieve a new one from the population.
At the left hand side of Figure \ref{fig:web}, the web page shows the \emph{Collections} section.
The user can create several collections, to group and organize his favorite 
artifacts. Moreover, users can browse the content of each collection and from
there share images through their social network. This makes the assignment 
of fitness through the rating system a {\em social} activity, 
pursuing the objective of this work, which is to increase engagement of the users.

>>>>

Some C-IEC systems promote user engagement by presenting interesting information to 
users, for instance the genetic lineage of each phenotype or the most popular or 
best rated solucion \cite{picbreeder,forms}. In a recent work by 
Wagy \& Bongard \cite{wagy2014collective} user interaction 
is needed for evaluating fitness and developing
new designs of robot locomotion. Collaboration is encouraged by gamifying the system 
using the maximum distance indicator to inspire the user to try and ``beat'' previous designs. 
In any case, using gamification techniques imply that we must deal with IEC
systems as socio-technical constructs, where the social aspects are
essential to understand its dynamics. In this sense, conclusions
reached with other systems such as NodEO \cite{DBLP:conf/gecco/MereloCGCRV16}
can also be applied to these systems; and applying social 
network techniques such as graph analysis
to their study will allow us to understand them more thoroughly. 
% C-IEAS
Given the human fatigue limitation when applying IEAs, some authors 
have tried to mitigate the problem by allowing the algorithm to 
collaborate with the user, so that sometimes 
users perform the evaluation,  but also specific measures are included 
into the algorithm to perform the
evaluation of some features automatically. For instance, 
Reis et al. added some terrain measures (such as accessibility and edge length) 
to a standard  IEC in \cite{DBLP:journals/soco/FradeVC12}. 
This way the algorithm was capable of providing terrains that would otherwise have needed the
users' evaluation for these specific features. Seyama and Munetomo \cite{seyama2016development}
also propose the reduction of user fatigue by using 
a collaborative filtering algorithm to show only the information utilized by similar users as 
they collaborate with a large number of users for the interactive modeling of 3D glasses. 
The framework and data model for 
representing the interactive process of our C-IEC systems is proposed next.
% Volunteer Based
% HBC
% Engagement Techniques
% Models

@@

\section{Human-Centered IEC Framework}
\label{sec:framework}
The general goal of this research is to develop a human-centered \cite{gasson2003human} 
software framework for C-IEC. 
A framework is defined as a reusable architectural design together
with an implementation \cite{campbell1991choices}, in this case 
providing generalized components to developers of IEC systems.

The proposed framework includes components that can be refined to increase
participation and also to minimize the amount of evaluations needed for the evolutionary 
process in a given IEC application. Software frameworks often have 
a vison \cite{carneiro2010introducing} guiding their
design. Before diving into details the main design considerations of the
framework are explained next:

\begin{itemize}
\item {\bf Users are human.} 
  The framework follows the approach of human centered computing \cite{sebe2010human},
  in which the context, environment, interfaces, preferences, accessibility, human relations,
  cognitive limitations, culture, creativity and other human aspects are an integral part 
  of the system. Humans are the computing resorces of the system, having unique characteristics
  as those identified by Sun \& Dance \cite{Sun2013}:
\begin{quote}
  (1) humans can solve computer hard problems; (2) humans are very good at exception handling,
  (3) humans have creativity, (4) humans have cognitive load limitation, (5) humans are
  vulnerable to psychological manipulation, (6) humans are prone to errors,
  especially for reflective tasks.
\end{quote}

\item {\bf Users are volunteers.} Users donate their computing resources, so they are 
unaccountable, sometimes they try to game the system. Project owners must actively promote and
design the interactive system to engage volunteers \cite{oh2015clicking}. % Define
\item {\bf Users are not alone}
  Relationship between users in an interactive evolutionary algorithm can be modeled
  as a social network, with well established semantics, algorithms and metrics 
  \cite{ahuja1993network}.
  A graph model could enable researchers to find other ways of identifying leaders of 
  opinion or measuring the similarity between user's preferences. 
  These measures can then be used by recommender algorithms selecting 
  phenotypes according users' preferences.

\item {\bf Context of use matters.}
  Fischer \cite{fischer2012context}
  defines context as
  \begin{quote}
  the interaction between humans and
  computers in socio-technical systems that takes place in a certain
  context referring to the physical and social situation in which
  computational devices and environments are embedded.
\end{quote}   
  Fischer also identifies the important aspects to consider when the context is used: how is
  contextual data obtained, how is context represented and what
  goals and purposes the context has when it is used in a particular
  application. An IEC system will  be used within a certain range 
  of technical, physical and social or
  organizational environments \cite{maguire2001context} that may affect its use.

\item {\bf Interaction is a stream of actions.}
  Real time processing of users' actions could be needed for certain applications when data is 
  captured by sensors, or directly captured as user input. For example, social networks keep
  track of users' interactions with other users, media objects and places. Users of 
  social networks (for instance the Facebook Graph) are accustomed to express these 
  complex relationships in sentences such as: ``John and Ann eating breakfast at Tony's''. 
  Other example is the W3C Activity Streams 2.0 specification used for representing activities 
  common in social web applications \cite{json:streams}. 
\end{itemize}

@@

\begin{enumerate}
  \item {\bf Interactive System.} 
  This is the real world system that we are going to represent in the data model, 
  it consist of human users and their interactions with one or more phenotypes
  from the population. There are many ways in which humans could interact 
  with phenotypes. The interaction consists of a set of actions and 
  takes place in a certain context. For example, through a mobile device  
  or by interacting with real world objects 
  \cite{de2014artists,de2013unplugging}. 
  There is also the possibility that fitness or even part of the search 
  is done by devices lent by humans \cite{DBLP:conf/gecco/MereloCGCRV16}.
  The information gathered through the interaction is the primary concern
  as it will guide the search.

  \item {\bf IEC Data Model.}
  A data model is used to describe the IEC system prior to a physical 
  implementation.  Depending on the domain several techniques can be used.
  Parts of the system could be better described by an entity-relationship 
  modeling, to be used in a relational database or a class digram for a 
  key-value store. A graph is proposed for modeling the social network of users 
  and their interaction with the population. When implemented, a graph database 
  will be the back-end of the system.

  \item {\bf Evolutionary Algorithm.} 
  The Evolutionary Algorithm (EA) algorithm interacts with the data model. The EA could also be run with the help of
  humans, for instance in the XY project \cite{de2013unplugging} artists actually painted
  the solutions of the new population and applied every genetic operator.
\end{enumerate}

@@

\begin{itemize}
\item {\bf EvoSpace-Redis.} 
EvoSpace is a population store \cite{Evospace}  for the development of 
evolutionary algorithms that are intended to run on a cloud computing model. 
The population is decoupled from any particular evolutionary algorithm. 
Candidate solutions are stored as of objects in a population, and they can be withdrawn, 
processed and replaced using a specified set of methods \cite{GValdez2015}. The population
is stored in-memory, using the Redis key-value database. Redis was chosen over a relational 
database management system because it provides a hash based implementation of sets and 
queues which are natural data structures for the EvoSpace model. Basicaly, a sample of 
candidate solutions are retrieved from the server, evaluated and then sent back. 
The same operations are used to evolve the population. In EvoSpace individuals replaced 
in the population are stored indefinitly, to permit users to store permalinks to them.
Implementation details are presented in \cite{garcia2013evospace}.

\item {\bf Graph-Neo4J.} 
 Collaborative IEC (C-IEC) systems need to store highly connected data, as it is common 
 in current applications like social networks. In order to deal with large datasets of connected 
 data found in these systems, graph databases \cite{angles2012comparison} have been proposed 
 as an alternative to relational databases which have performance limitations when dealing with 
 highly connected data \cite{holzschuher2013performance}.
 A graph is proposed for modeling the social network of users, their interaction with 
 candidate solutions, and the relationships between them in the population.
 The graph database system used in the implementation is Neo4J as it is
 a scalable solution \cite{miller2013graph,holzschuher2013performance}, and well 
 supported and documented in PaaS infrastructures like Heroku. The Cypher query 
 language it used to retrieve views from the graph.
 An example query is shown in Figure \ref{fig:cypher} where the relations 
 between users and solutions are presented.

@@

\item {\bf Relational Objects-PostgreSQL.}
  The PostgreSQL relational database system is also employed because user 
  sessions and authentication,
  as well as dynamic web pages are handled directly by the Django web
  framework \cite{garcia2013evospace}. % these implementations details
                                % could go if needed.

\item {\bf Interfaces.}
The appropiate human interface will depend on the application domain, 
the current framework implemantation employs a web based application. Developers 
can use different templates, to present one or more phenotypes 
at a time, and the type of rating system: like based or a rating from one to five (stars). 
Other implementations do not need a graphical interface at all, for example the
EvoDrawings instalation or the XYZ Project \cite{de2014artists}.

\item {\bf Evolution.}    
The EA is decoupled from the data model, the algorithm 
could be implemented as an EvoWorker \cite{garcia2013evospace}, use an external library, 
or even assigned to humans. An additional tool was implemented for a human based evolution, 
where volunteers select the parents of the next generation and upload the 
new individuals \cite{de2014artists}.  
\end{itemize}

\section{Case Study: EvoDrawings}
\label{sec:evodraw}
As a case study, a C-IEC application was developed using the 
EvoSpace-Interactive (ES-I) platform \cite{garcia2013evospace}
A brief description of the application is presented next, focusing
on the data elements that where ported to the graph model.

@@

\subsection{Fitness Assignment}
The ES-I platform has been programmed to employ a 
collaborative methodology for performing fitness assignment. 
Several registered users assign a quality assessment to a single
phenotype and then an aggregated fitness value is calculated,
depending on the rating (from 1 to 5 stars) by each particular user,  
resulting in a many-to-many relationship between users and
phenotypes. Other components or users of the systems can 
query this user-phenotype relation to extract relevant
knowledge about the process and the population. For instance, showing the
most popular phenotype, or the user with more participation 
\cite{picbreeder}.
In order to do this, metadata about phenotypes 
must be permanently stored, even
if they are no longer in the active population.

\subsection{Collaboration}
\label{sec:col}
Users need to authenticate themselves to the system using their
Facebook account. In that sense, some users might not be interested
either because they do not want to give that information or simply because
they do not use that social network. Even if we might lose
some users that way, the additional information we obtain for scoring
phenotypes more than compensates for that. 
After entering the web application by using their Facebook account,
users can collaborate with their Facebook friends, 
sharing those phenotypes they like, or by taking phenotypes
from their friend's collections by using the web interface depicted 
in Figure \ref{fig:web}.
At the top left corner a list of Facebook friends is presented
to encourage users to interact with the system. In the central 
\emph{ Wall } area, a phenotype sampled from the population that is
being evolved via the evolutionary algorithm 
is shown to the user.
Here, the user can interact with the system in two ways.
First, he can assign a rating to the phenotype or choose to add an image to one of their \emph{Collections}.
A collection is a special folder that stores those phenotypes a user likes and wishes
to save. After the user finishes interacting with the phenotype
on the Wall, he can choose to retrieve a new one from the population.
At the left hand side of Figure \ref{fig:web}, the web page shows the \emph{Collections} section.
The user can create several collections, to group and organize his favorite 
artifacts. Moreover, users can browse the content of each collection and from
there share images through their social network. This makes the assignment 
of fitness through the rating system a {\em social} activity, 
pursuing the objective of this work, which is to increase engagement of the users.
